Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2022-08-23 analysis/paper/paper-doc.qmd
split captions to landmarks out from schematic figure

![Korean Paleolithic sites mentioned in this study](images/site-map.png){#fig-map width="477"}

@@

In addition to attributes from previous studies of cultural transmission and projectile points, i.e. maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used additional morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-two). We obtained our morphological attribute data from landmark analysis of digitized images of the artifacts. Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life]. We did not include weight and thickness as these measurements were not available to us. For the landmark analysis, we put a total of 11 landmarks on the outline of each stemmed point and computed distances between pairs of landmark coordinates to derive attributes for our statistical analyses. The landmarks we recorded were: (1) the tip of the point; (2) the left side of mid-point; (3) the left wing, which can be overlapped with the mid-point; (4) the left side of the mid-point on the tang curve; (5) the left side of 5mm above the basal end; (6) the center of the basal end; (7) the tip of the basal end, which can be overlapped with the center of the basal end ; (8) the right side of 5mm above the basal end ; (9) the right side of the mid-point on the tang curve; (10) the right wing; and (11) the right side of mid-pint, which can be overlapped with the right wing. We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks and export them as XY coordinate data.

![(a) Stemmed point from Yongsandong site. with landmarks showing the attributes considered in this study and corresponding landmarks:\
ML: Maximum length, tip to bottom, perpendicular to the length axis 1- 7\
BL: Body length, tip to the closest wing, perpendicular to the length axis 1- 3 or 1-10 depending on the artifact\
TL: Tang length, the closest wing from the tip to bottom,perpendicular to the length axis 3 - 7 or 10-7 depending on the artifact\
SL: (Maximum) stem length, perpendicular to the length axis. Closer tang curve's middle point from the tip to the most distant point of the basal end 9-7 or 4-7 depending on the artifact\
MW: mid width, dimension from margin to margin at the mid-point of the length 2 - 11 axis, perpendicular to the width axis.\
TW: tang width, dimension between each wing, perpendicular to the width axis 3 - 10\
SW: stem width, width of the basal end of the point, 5mm above the end 5 - 8](images/ISKM%20talk%202021-cut.png){#fig-two width="221"}

@@

All data treatment, analyses and visualization were performed in the R environment[@core2021team]. Our R code and data including the raw images are openly available online: https://github.com/parkgayoung/CTtps as well at DOI to enable full reproducibility [@marwick2017computational].

@@

@fig-plot-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm). Among these attributes, stem width (SW) is the smallest and most highly standardized, regardless of overall size of the stemmed points. This result implies a similar thickness of a wooden shaft that is attached to the stem.

```{r}
#| label: fig-plot-attributes
#| fig-cap: Boxplot for each attribute measured on the stemmed points. The black bold horizontal lines in the middle of boxes represent the median value. The dots represent individual artefacts.
#| fig-width:   4.45
#| fig-height:  4.5

@@

# make box plot for each attributes
ggplot(data_box,
       aes(key, value)) +
  geom_boxplot() +
  geom_beeswarm(alpha = 0.3,
                size = 0.8,
                cex = 0.7) +
  ylab("Length (cm)") +
  xlab("Variables") +
  theme_bw()
```

@@

We used CV to measure the variations between stemmed points. If the variation is high, the transmission might happen through guided variation while the low variation represents indirect bias. Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes). Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher. Unlike the distribution of actual attributes (@fig-plot-attributes), the CVs of body length and maximum length have the lowest values. The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which could be the result of indirect bias.

```{r}
#| label: fig-plot-cv-all-attributes
#| fig-cap: The points and blue colored numbers represent CV values for each attribute measured on the stemmed points. 95% confidence intervals for the CVs were computed using Sharma & Krishna's method. The vertical lines indicate the range of confidence intervals.
#| fig-width:   4.8
#| fig-height:  4.5

ggplot(cv_data_sharma_named,
       aes(attribute,
           all_cv_sharma)) +        # ggplot2 plot with confidence intervals
  geom_point() +
  geom_errorbar(aes(ymin = all_low_sharma, 
                    ymax = all_high_sharma), 
                width = 0.2) +
  ylab("CV") +
  xlab("Variables") +
  geom_text(aes(label = round(all_cv_sharma,1)), 
            col="blue", 
            hjust = -0.3, 
            size = 3) +
  theme_bw()
```

@@

PCA serves to examine correlation of attributes for determining the dominant mode of cultural transmission. @fig-five shows positive relationship between attributes. Especially, there are strong relationships between body length and maximum length, tang width and mid width, and tang length and stem length. Body length and maximum length contribute the most to the shape of stemmed points and they are most correlated to each other. Other width and tang related attributes are close to each other but a bit far from body length and maximum length. Stem width contributes the least to the overall shape. We can assume that the body length and maximum length were prioritized when the stemmed points were transmitted.

![PCA analysis for stemmed points. The direction of the arrow shows the relationship between attributes. The length and color of the arrow indicate the strength of contribution to the stemmed point shape.](images/007-PCA-1.png){#fig-five width="284"}

4.  Variation Over Time

@@

Compared to Phase 2, the influence of tang length and stem length of Phase 3 becomes more pronounced and while stem width is less effective (@fig-seven). However, there are no notable changes in correlations over time.

![PCA analysis for the second and third chronological phases in the Korean Late Paleolithic period. The direction of the arrow shows the relationship between attributes. The length and color of the arrow indicate the strength of contribution to the stemmed point shape.](images/007-PCA.png){#fig-seven width="534"}

@@

5.  Variation Between sites

@@

cv_by_full_site_df_label <-
  df_full_sitename %>%
  dplyr::select(-SPstage.Stage) %>%
  group_by(full_sitename) %>%
  add_tally() %>%
  filter( n > 1) %>%
  mutate(label = as.factor(paste0(full_sitename,' (N = ', n, ")"))) %>%
  dplyr::select(-n) %>%
  nest(-full_sitename, -label) %>%
  mutate(cv_by_site = map(data, ~map_df(.x, cv))) %>%
  dplyr::select (-cv_by_site)

@@

cv_plot_site_label <- cv_by_site_df %>%
  left_join (cv_by_full_site_df_label) %>%
  dplyr::select (- data,-full_sitename)

@@

6.  Variation Between Raw Materials

We examined the relationship between CV values and raw materials with a premise that the shape of stemmed points are highly dependent on raw materials. We exclude raw materials that were used for single point. @fig-ten shows that quartzite, the raw materials for Sachang stemmed points, has the highest CV value with a wide range of conference intervals. Quartz has high variations of CV among the attributes. On the other hand, shale has the lower CV, which is very stable throughout the attributes.

![CV values for raw materials that were used for making more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.](images/00000-cv-raw-materials.png){#fig-ten width="526"}

>>>>

```{r}
#| label: fig-map
#| fig-cap: Korean Paleolithic sites mentioned in this study.
#| fig-width:   4.45
#| fig-height:  4.5

# Thisdraws the map

korean_archaeological_site_locations <-
  readr::read_csv(here::here("analysis/data/raw_data/sitelocation.csv"))

korean_archaeological_site_locations <-
  tibble::rowid_to_column(korean_archaeological_site_locations, "ID")

library(ggmap)
library(ggrepel)
# devtools::install_github('3wen/legendMap')
library(legendMap)

# load map tiles we previously saved, we got these using the code in 
# 000-download-map-tiles.R
map <- readRDS(here::here("analysis/data/raw_data/map_tiles.rds"))

# map with site names next to the points
map_of_sites <-
  ggmap(map)  +
  geom_point(data = korean_archaeological_site_locations,
             aes(long_dd ,
                 lat_dd),
             colour = "red",
             size = 2) +
  # label points with site name
  geom_text_repel(data = korean_archaeological_site_locations,
                  aes(long_dd ,
                      lat_dd,
                      label = full_sitename),
                  size = 2,
                  bg.color = "white",
                  bg.r = 0.15) +
  theme_minimal(base_size = 6) +
  labs(x = "Longitute",
       y = "Latitude") +
  legendMap::scale_bar(
    # edit these numbers to select a suitable location
    # for the scale bar where it does not cover
    # important details on the map
    lon = 124.5,
    lat = 33.5,
    legend_size = 2,
    # distance of one section of scale bar, in km
    distance_lon = 50,
    # height of the scale bar, in km
    distance_lat = 5,
    # distance between scale bar and units, in km
    distance_legend = 20,
    # units of scale bar
    dist_unit = "km",
    # add the north arrow
    orientation = TRUE,
    # length of N arrow, in km
    arrow_length = 60,
    # distance between scale bar & base of N arrow, in km
    arrow_distance = 50,
    # size of letter 'N' on N arrow, in km
    arrow_north_size = 10) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank())

# inset world map, from
# https://stackoverflow.com/a/62104396/1036500
library(maps)
library(magrittr)
library(maptools)
library(raster)

#Defining a general CRS
mycrs <- "+proj=longlat +datum=WGS84 +no_defs"

#Using the original maps package, then converting map into SpatialPolygons object
world <- maps::map("world", fill=TRUE,  plot = FALSE) %$%
  maptools::map2SpatialPolygons(., IDs=names,proj4string=CRS(mycrs))

#The resulting map has self intersection problems so any further operation reports errors; using buffers of width 0 is a fast fix
while(rgeos::gIsValid(world)==FALSE){
  world <- rgeos::gBuffer(world, byid = TRUE, width = 0, quadsegs = 5, capStyle = "ROUND")
}

#Dissolving polygon's limits
world <- raster::aggregate(world)

#Plotting. I add theme_void to your code to erase any axis, etc
worldplot <-
  ggplot() +
  geom_polygon(data = world,
               aes(x=long, y=lat,
                   group=group),
               fill='NA',
               color='black',
               size=0.1) +
  annotate("rect",
           xmin = 118,
           xmax = 137,
           ymin = 28,
           ymax = 47,
           fill = NA,
           colour = "red",
           size = 0.5
  ) +
  theme_void() +
  coord_fixed(1) +
  labs(x = "", y = "") +
  labs(x = NULL, y = NULL, title = NULL) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.border = element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

# combine site map and inset map
library(cowplot)
ggdraw() +
  draw_plot(map_of_sites) +
  draw_plot(plot = worldplot,
            x = 0.6, # x location of inset placement
            y = -0.09, # y location of inset placement
            width = .4, # Inset width
            height = .35, # Inset height
            scale = 0.5 # Inset scale
  ) +
  theme(panel.background = element_rect(fill='white', colour="white"),
        plot.background = element_rect(fill='white', colour="white"))
```

@@

In addition to attributes from previous studies of cultural transmission and projectile points, i.e. maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used additional morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic). We obtained our morphological attribute data from landmark analysis of digitized images of the artifacts. Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life]. We did not include weight and thickness as these measurements were not available to us. For the landmark analysis, we put a total of 11 landmarks on the outline of each stemmed point and computed distances between pairs of landmark coordinates to derive attributes for our statistical analyses. The landmarks we recorded were: (1) the tip of the point; (2) the left side of mid-point; (3) the left wing, which can be overlapped with the mid-point; (4) the left side of the mid-point on the tang curve; (5) the left side of 5mm above the basal end; (6) the center of the basal end; (7) the tip of the basal end, which can be overlapped with the center of the basal end ; (8) the right side of 5mm above the basal end ; (9) the right side of the mid-point on the tang curve; (10) the right wing; and (11) the right side of mid-pint, which can be overlapped with the right wing. We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks and export them as XY coordinate data.

```{r}
#| label: fig-labelled-schematic
#| fig-height: 2
#| fig-cap: Stemmed point from Yongsandong site. with landmarks showing the attributes considered in this study and corresponding landmarks

knitr::include_graphics(here::here("analysis/figures/hand-made-figures/ISKM-talk-2021-cut.png"))
```

```{r}
#| label: tab-landmark-abbreviations

here::here("analysis/data/raw_data/table-of-landmark-abbreviations.csv") %>% 
  read_csv() %>% 
  knitr::kable(., caption = "Description of artefact landmarks used in this study")

@@

All data treatment, analyses and visualization were performed in the R environment [@core2021team]. Our R code and data including the raw images are openly available online: https://github.com/parkgayoung/CTtps as well at DOI to enable full reproducibility [@marwick2017computational].

@@

@fig-plot-cv-all-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm). Among these attributes, stem width (SW) is the smallest and most highly standardized, regardless of overall size of the stemmed points. This result implies a similar thickness of a wooden shaft that is attached to the stem.

```{r}
#| label: fig-plot-attributes

@@

# make box plot for each attributes
plot_raw_measurements_for_attributes <- 
ggplot(data_box,
       aes(key, value)) +
  geom_boxplot() +
  geom_beeswarm(alpha = 0.3,
                size = 0.8,
                cex = 0.7) +
  ylab("Length (cm)") +
  xlab("") +
  theme_bw()
```

@@

We used CV to measure the variations between stemmed points. If the variation is high, the transmission might happen through guided variation while the low variation represents indirect bias. Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes). Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher. Unlike the distribution of actual attributes (@fig-plot-cv-all-attributes panel A), the CVs of body length and maximum length have the lowest values. The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which could be the result of indirect bias.

(ref:cap-fig-plot-cv-all-attributes) A: Boxplot for each attribute measured on the stemmed points. The black bold horizontal lines in the middle of boxes represent the median value. The dots represent individual artefacts. B: The points and blue colored numbers represent CV values for each attribute measured on the stemmed points. 95 p confidence intervals for the CVs were computed using Sharma and Krishna's method. The vertical lines indicate the range of confidence intervals.

```{r}
#| label: fig-plot-cv-all-attributes
#| fig-cap: '(ref:cap-fig-plot-cv-all-attributes)'
#| fig-width:   7
#| fig-height:  4

plot_cv_all_attributes <- 
ggplot(cv_data_sharma_named,
       aes(attribute,
           all_cv_sharma)) +        # ggplot2 plot with confidence intervals
  geom_point() +
  geom_errorbar(aes(ymin = all_low_sharma, 
                    ymax = all_high_sharma), 
                width = 0.2) +
  ylab("CV") +
  xlab("Variable") +
  geom_text(aes(label = round(all_cv_sharma,1)), 
            col="blue", 
            hjust = -0.3, 
            size = 3) +
  theme_bw()

library(cowplot)
plot_grid(plot_raw_measurements_for_attributes,
          plot_cv_all_attributes,
          ncol = 1, 
          labels = "AUTO")
```

@@

PCA serves to examine correlation of attributes for determining the dominant mode of cultural transmission. @fig-pca-attributes shows positive relationship between attributes. Especially, there are strong relationships between body length and maximum length, tang width and mid width, and tang length and stem length. Body length and maximum length contribute the most to the shape of stemmed points and they are most correlated to each other. Other width and tang related attributes are close to each other but a bit far from body length and maximum length. Stem width contributes the least to the overall shape. We can assume that the body length and maximum length were prioritized when the stemmed points were transmitted.

```{r}
#| label: fig-pca-attributes
#| fig-cap: PCA analysis for stemmed points, for all phases combined, then for only phase two and phase three. The direction of the arrow shows the relationship between attributes; arrows pointing in the same direction indicate variables that are highly postitively correlated. The length of the arrow indicates the strength of contribution to the stemmed point shape.
#| fig-width:   7
#| fig-height:  3

library(ggbiplot)
library(ggfortify)
### PCA analysis using ggbiplot, ggfortify pkgs

PCA <- prcomp(DF, center = TRUE, scale = TRUE)

# remove stage 1 artefact
xx_pca <-
  df_sitename %>%
  dplyr::filter(SPstage.Stage != 1) %>%
  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
  prcomp(., center = TRUE, scale = TRUE)

xx_pca1 <-
  data.frame(xx_pca$x,
             stage = as.factor( df_sitename$SPstage.Stage[df_sitename$SPstage.Stage !=1]))

## CV with site (n=X)
label_phase <-
  df_sitename %>%
  group_by(SPstage.Stage) %>%
  add_tally() %>%
  mutate(label = as.factor(paste0(SPstage.Stage,'\nN = ',n)))

count_phase <- length(which(label_phase[9]>0))
count_phase2 <- length(which(label_phase[9] == 2))
count_phase3 <- length(which(label_phase[9] == 3))

# phase 2

xx_pca2 <-
  df_sitename %>%
  dplyr::filter(SPstage.Stage == 2) %>%
  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
  prcomp(., center = TRUE, scale = TRUE)

# phase 3

xx_pca3 <-
  df_sitename %>%
  dplyr::filter(SPstage.Stage == 3) %>%
  dplyr::select(-sitename, -SPstage.Stage, -SPstage.Raw_material) %>%
  prcomp(., center = TRUE, scale = TRUE)

##### using factoextra pkg
library(factoextra) #http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/

# Color variables by the continuous variable
cv_all <-
fviz_pca_var(xx_pca) +
  ggtitle(paste0("All phases", " (n = ", count_phase,")")) +
  theme(legend.position = "none")

# phase 2
cv_two <-
  fviz_pca_var(xx_pca2) +
  ggtitle(paste0("Phase 2", " (n = ", count_phase2,")"))

# phase 3
cv_three <-
  fviz_pca_var(xx_pca3) +
  ggtitle(paste0("Phase 3", " (n = ", count_phase3,")"))

library(patchwork)
cv_all + cv_two + cv_three
```

## Variation Over Time

@@

Compared to Phase 2, the influence of tang length and stem length of Phase 3 becomes more pronounced and while stem width is less effective (@fig-pca-attributes). However, there are no notable changes in correlations over time.

@@

## Variation Between sites

@@

cv_by_full_site_df_label <-
  df_full_sitename %>%
  dplyr::select(-SPstage.Stage) %>%
  group_by(full_sitename) %>%
  add_tally() %>%
  dplyr::filter( n > 1) %>%
  mutate(label = as.factor(paste0(full_sitename,' (N = ', n, ")"))) %>%
  dplyr::select(-n) %>%
  nest(-full_sitename, -label) %>%
  dplyr::mutate(cv_by_site = purrr::map(data, ~map_df(.x, cv))) %>%
  dplyr::select(-cv_by_site)

@@

cv_plot_site_label <- cv_by_site_df %>%
  left_join(cv_by_full_site_df_label) %>%
  dplyr::select(- data,-full_sitename)

@@

## Variation Between Raw Materials

We examined the relationship between CV values and raw materials with a premise that the shape of stemmed points are highly dependent on raw materials. We exclude raw materials that were used for single point. @fig-plot-cv-by-raw-material shows that quartzite, the raw materials for Sachang stemmed points, has the highest CV value with a wide range of conference intervals. Quartz has high variations of CV among the attributes. On the other hand, shale has the lower CV, which is very stable throughout the attributes.

```{r}
#| label: fig-plot-cv-by-raw-material
#| fig-cap: CV values for raw materials that were used for making more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.
#| fig-width:   8.5
#| fig-height:  4.5

# CV by raw materials
### Make a table of CVs for all variable grouped by raw materials
cv_by_raw_materials <-
  df_full_sitename %>%
  dplyr::select(-SPstage.Stage, -lat_dd, -long_dd, -sitename, -full_sitename) %>%
  group_by(SPstage.Raw_material) %>%
  add_tally() %>%
  dplyr::filter( n > 1) %>%
  dplyr::select(-n) %>%
  pivot_longer(-SPstage.Raw_material,
               names_to = "variable",
               values_to = "value") %>%
  group_by(SPstage.Raw_material, variable) %>%
  dplyr::summarise(cv_by_materials =   sharma_cv(value),
            cv_low_sharma =  sharma_int_low(value),
            cv_high_sharma = sharma_int_high(value))

######CV for per raw materials (n=X) with the full site name : label
cv_by_raw_materials_df_label <-
  df_sitename %>%
  group_by(SPstage.Raw_material) %>%
  add_tally() %>%
  dplyr::filter( n > 1) %>%
  mutate(label = as.factor(paste0(SPstage.Raw_material,' (N = ', n, ")"))) %>%
  dplyr::select(-n) %>%
  nest(-SPstage.Raw_material, -label) %>%
  mutate(cv_by_site = purrr::map(data, ~map_df(.x, cv))) %>%
  dplyr::select (-cv_by_site)

######add label to main dataframe
cv_plot_raw_materials_label <- cv_by_raw_materials %>%
  left_join (cv_by_raw_materials_df_label) %>%
  dplyr::select (- data,-SPstage.Raw_material)

###### facet plot
cv_plot_raw_materials_label %>%
  ggplot(aes(x = variable, y = cv_by_materials)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma), width = .2) +
  xlab("Law Materials") +
  ylab("Coefficient of Variation on Attributes") +
  facet_wrap( ~ label, scales = "free_y") +
  coord_cartesian(ylim = c(0,130)) +
  theme_bw() 
```
